Investigation of the Effect of Pretraining on EEG Classification with Deep Recurrent-Convolutional Networks

### 2017-10-15 paper/paper.tex
removed parameter column from tables
scratched a sentence
unbolded a claim
fix grammar/style issues

\section{Introduction}
\subparagraph{Motivation} Over the last decades, neural networks have become an extremely popular approach for the data analysis in various fields. An artificial neural network (ANN) is a computational model that is designed to simulate the way the human brain analyzes and processes information. ANNs are based on the structure and functions of biological neural networks and are able to automatically learn and improve from experience without being explicitly programmed. ANNs are considered nonlinear statistical data modeling tools where the complex relationships between inputs and outputs are modeled or patterns are found.

Machine learning (application of ANNs) techniques have proven to play a crucial role in many application areas, because they possess a high classification power, are able to compute high-level abstractions of data, and provide insights into hidden relationships among the data. This is especially important in the field of neuroscience, medical signal and image processing. Often data sets might have incomplete information, which is compounded by a poor understanding of the physical mechanisms by which the data is generated and subjective evaluation of the data by a human observer \cite{roberts}. Electroencephalography (EEG) data analysis relies a lot on statistics, and one of the challenges in modeling cognitive events from EEG data is finding representations that are invariant to inter- and intra-subject differences, therefore, machine learning methods are a crucial tool for EEG-based research. There are many clinical medical applications which include EEG-based brain-computer interface systems, or devices that are intended to change the lives of patients with disabilities, and a lot of research is being done in the area.

\subparagraph{Related work} Many recent studies \cite{cecotti, shamwell,driver} investigated the application of convolutional neural networks (ConvNets), which are neural networks that can learn local patterns in data by using convolutions as their key component, in the field of EEG analysis. In mathematics convolution is an operation on two functions to produce a third function. Each convolutional layer provides a more abstract feature representation, applying a convolution operation to the input, passing the result to the next layer. Convolution here means same filter applied in parallel to all positions in the input. Convolutional layer applies multiple filters to all input positions from layer below.

Schirrmeister et al. \cite{schirrmeister} showed how to design and train deep ConvNets with a range of different architectures to achieve the prominent results in EEG decoding tasks by applying recent advances from the machine learning field and focusing on learning from the raw data of EEG. The architecture desing Schirrmeister et al. used was a  deep ConvNet with four convolution-max-pooling blocks, with a special first block designed to handle EEG input, followed by three standard convolution-max-pooling blocks and a dense softmax classification layer. The input was represented as a 2D-array with the number of time steps as the width and the number of electrodes as the height. This approach significantly reduced the input dimensionality compared with another approach to represent EEG input as a time series of topographically organized images, which was done by Bashivan et al. \cite{learning_eeg_repr}.

Sleep analysis still remains the most unexplored area of neuroscience. One of the bottlenecks in sleep research is the analysis of the EEG data. Many steps, such as identifying the stages of sleep, require visual inspection of the data. Performed by human scientists, scoring and analysis are slow and biased. Neural networks have been successfully applied to identify different sleep stages and their applications for extracting categorical cognitive sleep information are a huge focus of on-going research. Laengvist et al. \cite{sleep} proposed the use of an architecture called deep belief nets (DBNs). A deep belief network is a probabilistic generative model with deep architecture that searches the parameter space by unsupervised greedy layerwise training. Laengvist et al. applied DBNs to the raw EEG input and also validated learned representations by integrating a hidden Markov model to capture a more realistic sleep stage switching, so that the knowledge from the human experts is integrated to the system.

@@

The approach to analysing and modeling EEG during sleep recordings, that we based our project on, was presented by Bashivan et al. in the paper "Learning Representations from EEG with Deep Recurrent-Convolutional Neural Networks" \cite{learning_eeg_repr}. Bashivan et al. used a novel approach, transforming EEG data into a sequence of multi-spectral images which preserve spatial information, which, in its turn, is an important property of EEG methodology. The proposed approach showed significant improvements in accuracy of the performance on the cognitive load classification task.

\section{Data}
\subsection{Data origin \& structure}
\label{ssec:data_structure}
The data set was created at the Sleep Research Laboratory (Prof. Stuart Fogel, PhD) at the University of Ottawa in Canada and kindly shared with us for the purpose of this project. The researchers in the laboratory analyse and model brain activity, with the use of Electoencephalography (EEG). EEG is a noninvasive method to record electrical activity of the brain, by placing electrodes along the scalp. We received data from one experiment which was conducted in the laboratory and focused on modeling dream and sleep activity and correlating it with sleep reports. The participants were asked to come to the lab, perform a virtual tennis game, or spatial navigation task, or both and then, directly after that, to go to sleep. After a short period of time, the participants were waken up and asked to verbally describe what they dreamt about. In some session types, the full night sleeps were additionally recorded. For this particular experiment, 22-channel EEG setup was applied, with the sampling rate of 128 samples per second, and the electrodes were customized to the participants according to the International 10-20 system, as can be seen in the figure 1.
\begin{figure}[!ht]
	\centering
	\includegraphics[scale=0.1]{eeg}
	\caption{International 10-20 EEG placement system}
\end{figure}

All of the EEG data we received was already cleaned, so no additional filtering, or artifact rejection was needed to be performed. An example of the EEG data before preprocessing can be seen in figure 2.  For 26 subjects who participated in the experiment there were 766 trials in total, and the experiment consisted of two parts: dream reports and mental rehearsals. Dream reports contained EEG data during the naps, mental rehearsals contained EEG recordings of when the participants were awake and asked to repeat the same task in their minds (to think about doing the task without actually doing it). Verbal reports were recorded for both of above mentioned parts and consisted of texts with the description provided by the participants. Each instance in the data was annotated with a label of the session type (tennis, spatial navigation, or both).
\begin{figure}[!ht]
	\centering
	\includegraphics[scale=0.2]{eeg_ex}
	\caption{The sample of EEG data from dream reports from one of the participants. EEG traces (horizontal: time in seconds; vertical: channel locations).}
\end{figure}

@@

\subparagraph{Motivation} EEG measurements are in form of 2-dimensional time series, indicating the change of electrode voltage over time. Such data in raw form contains noise, and has a relatively high dimensionality. Usually such data is analyzed in the frequency domain. This reduces significantly the size of the samples, and makes the influence of noise smaller. We have employed a data transformation method proposed by Bashivan et al.\cite{learning_eeg_repr}, which preserves the spatial information contained in the original data - the location of the electrodes. Measurements are aggregated over smaller periods of time, to form images of activity over the surface of the scalp. Different frequency bands are portrayed in different channels of the images. A data sample is then represented as a sequence of images. One of the motivations behind transforming the data to such a format was the fact, that there exist many Neural Network architectures successful in processing image- and video-based data. Data in visual format is also easier to analyze, what might also be helpful when investigating whether the model has learned useful features, eg. by introspection of Convolutional Filters (see Figure \ref{fig:inputs_optimized}).

@@

\begin{enumerate}
 \item Raw sequences shorter than the longest one were padded with zeros.
 \item Values were then (optionally) normalized by dividing by the highest found value.
\end{enumerate}
Data was then transformed into the frequency domain. For each trial:
\begin{enumerate}[resume]
 \item Fast Fourier Transform has been applied on fixed time windows of 1 second, with 0.5 second overlap 		with the previous window.
 \item The data has then been separated into the three frequency bands - Theta (4-8 Hz), Alpha (8-13 Hz) and Beta (13-30 Hz).
 \item For each band, sum of squared values have been computed
\end{enumerate}
Finally, transformation to images has been performed. For each time window:
\begin{enumerate}[resume]
 \item In order to project the data to images, oordinates of the electrodes have been transformed from 3D into 2D space using Azimuthal Equidistant Projection, a method used for geospatial purposes \cite{snyder}.
 The motivation behind this is that the shape of the head, on which the electrodes are placed, is similar in shape to a sphere.
 \item For each band, its corresponding image has then been constructed, using the newly obtained 2D points as pixel locations. Other pixel values have been interpolated with Clough-Tocher scheme \cite{alfeld}.
 \item An RGB image has been constructed by using the bands' images as channel data.  
\end{enumerate}

@@

\section{Methods}
\subsection{Baseline Methods}
Various non-deep learning methods have been employed for obtaining baseline results.
As in Bashivan et al. \cite[pp. 7-8]{learning_eeg_repr}, Support Vector Machines, Random Forest \cite{random_forests}, and sparse Logistic Regression models have been trained, using multiple parameter setting.
\subsubsection{Data Preparation}
For the baseline methods, the data has been preprocessed in a more simplified way than for the Neural Network models. This is due to the fact that the models used do not scale as well for high dimensional data, as opposed to the Convolutional Layers in the Neural Network models. Features are thus computed for the whole sequence, rather than extracting them per certain time steps. This results in data that is no more sequential in nature, but describes the whole observation.
For each of the 19 channels available in the data, Fast Fourier Transform has been applied on the whole length of a channel's data. Magnitude of each band (theta, alpha, and beta) has then been extracted and stored as the channel's features. In addition, an average over all channels has been computed. The resulting feature vector is then of size 60: \textit{(19 channels + additional averaged one) $\times$ 3 frequency bands}.
\subsubsection{Models}

\subparagraph{Support Vector Machines}
A Support Vector Machine classifier constructs a hyperplane (or set of) in the data space, which should separate the different classes. A model has been trained for each combination of the following parameters:
\begin{itemize}
	\item The penalty parameter \textbf{C} - from range of \{0.01, 0.1, 1, 10, 100\}
	\item The $\gamma$ kernel coefficient - from range of \{0.1, 0.2, .., 1, 2, .., 10\}
\end{itemize}

\subparagraph{Random Forests}
An ensemble classification method. A number of Decision Trees is being fit on different subsets of features of the training data. When classifying a data sample, the class which is given the majority of votes among the Decision Trees is chosen. A model has been trained for each of the following numbers of Decision Trees: 5, 10, 20, 50, 100, 500, 1000.

\subparagraph{Logistic Regression}
A model has been trained, using the $l_1$ as the norm for penalization, for each of the following penalization parameters \textbf{C}: \{0.01, 0.1, 1, 10, 100, 1000\}.

@@

The Neural Network models were trained on mini batches of size 32, using Adam \cite{Adam} as the optimization method.

@@

\begin{tabular}{ c|c|c } 
 Model & No. parameters & Accuracy \\ 

@@

 SVM & - & 7.8\% $\pm \, 4.1\% $\\
 Random Forest & - & 36.7\% $\pm \, 7.3\%$ \\
 Logistic Regression & - & 33.6\% $\pm \, 5.8\%$ \\

@@

 2 layer unidirectional LSTM & - & 18.6\% $\pm \, 2.0\% $\\
 2 layer bidirectional LSTM & - & 83.7\% $\pm \, 4.1\%$ \\
 3 layer bidirectional LSTM & - & 83.4\% $\pm \, 3.0\%$ \\
 5 layer bidirectional LSTM & - & 75.8\% $\pm \, 5.6\%$ \\
 2 layer bidirectional LSTM with 0.1 dropout & - & 84.1\% $\pm \, 4.2\%$ \\
 2 layer bidirectional LSTM with 0.2 dropout & - & 79.9\% $\pm \, 5.0\%$ \\

@@

 2 layer bidirectional LSTM with 0.1 dropout & - & 89.2\% $\pm \, 4.4\%$ \\

>>>>

\section{Introduction}
\subparagraph{Motivation} Over the last decades, neural networks have become an extremely popular approach for data analysis in various fields. An artificial neural network (ANN) is a computational model that is designed to simulate the way the human brain analyzes and processes information. ANNs are based on the structure and functions of biological neural networks and are able to automatically learn and improve from experience without being explicitly programmed. ANNs are considered nonlinear statistical data modeling tools where the complex relationships between inputs and outputs are modeled or patterns are found.

Machine learning (application of ANNs) techniques have proven to play a crucial role in many application areas, because they possess a high classification power, are able to compute high-level abstractions of data, and provide insights into hidden relationships among the data. This is especially important in the field of neuroscience, medical signal and image processing. Often datasets might have incomplete information, which is compounded by a poor understanding of the physical mechanisms by which the data is generated and subjective evaluation of the data by a human observer \cite{roberts}. Electroencephalography (EEG) data analysis relies heavily on statistics, and one of the challenges in modeling cognitive events from EEG data is finding representations that are invariant to inter- and intra-subject differences; therefore, machine learning methods are a crucial tool for EEG-based research. There are many clinical medical applications which include EEG-based brain-computer interface systems, or devices that are intended to change the lives of patients with disabilities, and a lot of research is being done in the area.

\subparagraph{Related work} Many recent studies \cite{cecotti, shamwell,driver} investigated the application of convolutional neural networks (ConvNets), which are neural networks that can learn local patterns in data by using convolutions as their key component, in the field of EEG analysis. Each convolutional layer provides a more abstract feature representation by applying a convolution operation to the input and passing the result to the next layer. Convolution here means the same filter is applied in parallel to all positions in the input. A convolutional layer applies multiple filters to all input positions from the layer below.

Schirrmeister et al. \cite{schirrmeister} showed how to design and train deep ConvNets with a range of different architectures to achieve the prominent results in EEG decoding tasks by applying recent advances from the machine learning field and focusing on learning from raw EEG data. The architecture design Schirrmeister et al. used was a  deep ConvNet with four convolution-max-pooling blocks, with a special first block designed to handle EEG input, followed by three standard convolution-max-pooling blocks and a dense softmax classification layer. The input was represented as a 2D-array with the number of time steps as the width and the number of electrodes as the height. This approach significantly reduced the input dimensionality compared with another approach of representing EEG input as a time series of topographically organized images, which was suggested by Bashivan et al. \cite{learning_eeg_repr}.

Sleep analysis remains one of the less explored areas of neuroscience. One of the bottlenecks in sleep research is the analysis of the EEG data. Many steps, such as identifying the stages of sleep, require visual inspection of the data. Performed by human scientists, scoring and analysis are slow and biased. Neural networks have been successfully applied to identify different sleep stages and their applications for extracting categorical cognitive sleep information are a huge focus of ongoing research. Laengvist et al. \cite{sleep} proposed the use of an architecture called deep belief nets (DBNs). A deep belief network is a probabilistic generative model with deep architecture that searches the parameter space by unsupervised greedy layerwise training. Laengvist et al. applied DBNs to the raw EEG input and also validated learned representations by integrating a hidden Markov model to capture a more realistic sleep stage switching, so that knowledge from the human experts is integrated to the system.

@@

The approach to analysing and modeling EEG during sleep recordings, on which we based our project, was presented by Bashivan et al. in the paper "Learning Representations from EEG with Deep Recurrent-Convolutional Neural Networks" \cite{learning_eeg_repr}. Bashivan et al. used a novel approach, transforming EEG data into a sequence of multi-spectral images which preserve spatial information, which, in turn, is an important property of EEG methodology. The proposed approach showed significant improvements in accuracy of the performance on the cognitive load classification task.

\section{Data}
\subsection{Data origin \& structure}
\label{ssec:data_structure}
The dataset was created at the Sleep Research Laboratory (Prof. Stuart Fogel, PhD) at the University of Ottawa in Canada and kindly shared with us for the purpose of this project. The researchers in the laboratory analyse and model brain activity, with the use of Electoencephalography (EEG). EEG is a noninvasive method to record electrical activity of the brain by placing electrodes along the scalp. We received data from one experiment which was conducted in the laboratory and focused on modeling dream and sleep activity and correlating it with sleep reports. The participants were asked to come to the lab, perform a virtual tennis game, or spatial navigation task, or both and then, directly after that, to go to sleep. After a short period of time, the participants were awakened and asked to verbally describe what they dreamt about. In some session types, the full night sleeps were additionally recorded. For this particular experiment, 22-channel EEG setup was applied, with a sampling rate of 128 samples per second, and the electrodes were customized to the participants according to the International 10-20 system, as can be seen in the figure 1.
\begin{figure}[!ht]
	\centering
	\includegraphics[scale=0.1]{eeg}
	\caption{International 10-20 EEG placement system}
\end{figure}

All of the EEG data we received was already cleaned, so no additional filtering or artifact rejection was needed to be performed. An example of the EEG data before preprocessing can be seen in figure 2.  For 26 subjects who participated in the experiment there were 766 trials in total, and the experiment consisted of two parts: dream reports and mental rehearsals. Dream reports contained EEG data during the naps, mental rehearsals contained EEG recordings of when the participants were awake and asked to repeat the same task in their minds (to think about doing the task without actually doing it). Verbal reports were recorded for both of above mentioned parts and consisted of texts with the description provided by the participants. Each instance in the data was annotated with a label of the session type (tennis, spatial navigation, or both).
\begin{figure}[!ht]
	\centering
	\includegraphics[scale=0.2]{eeg_ex}
	\caption{The sample of EEG data from dream reports from one of the participants. EEG traces (horizontal: time in seconds; vertical: channel locations).}
\end{figure}

@@

\subparagraph{Motivation} EEG measurements are in the format of a 2-dimensional time series, indicating the change of electrode voltage over time. Such data in raw form contains noise, and has a relatively high dimensionality. Usually such data is analyzed in the frequency domain. This reduces significantly the size of the samples, and makes the influence of noise smaller. We have employed a data transformation method proposed by Bashivan et al.\cite{learning_eeg_repr}, which preserves the spatial information contained in the original data - the location of the electrodes. Measurements are aggregated over smaller periods of time, to form images of activity over the surface of the scalp. Different frequency bands are portrayed in different channels of the images. A data sample is then represented as a sequence of images. One of the motivations behind transforming the data to such a format was the fact that there exist many Neural Network architectures successful in processing image- and video-based data. Data in visual format is also easier to analyze, which might also be helpful when investigating whether the model has learned useful features, eg. by introspection of Convolutional Filters (see Figure \ref{fig:inputs_optimized}).

@@

\begin{enumerate}
 \item Raw sequences shorter than the longest one were padded with zeros.
 \item Values were then (optionally) normalized by dividing by the highest found value.
\end{enumerate}
Data was then transformed into the frequency domain. For each trial:
\begin{enumerate}[resume]
 \item Fast Fourier Transform was applied on fixed time windows of 1 second, with 0.5 second overlap 		with the previous window.
 \item The data was then separated into the three frequency bands - Theta (4-8 Hz), Alpha (8-13 Hz) and Beta (13-30 Hz).
 \item For each band, sum of squared values were computed
\end{enumerate}
Finally, the transformation to images was performed. For each time window:
\begin{enumerate}[resume]
 \item In order to project the data to images, the coordinates of the electrodes were transformed from 3D into 2D space using Azimuthal Equidistant Projection, a method used for geospatial purposes \cite{snyder}.
 The motivation behind this is that the shape of the head, on which the electrodes are placed, is similar in shape to a sphere.
 \item For each band, its corresponding image was constructed, using the newly obtained 2D points as pixel locations. Other pixel values were interpolated with Clough-Tocher scheme \cite{alfeld}.
 \item An RGB image was constructed by using the bands' images as channel data.  
\end{enumerate}

@@

\section{Methods}
\subsection{Baseline Methods}
Various non-deep learning methods were employed for obtaining baseline results.
As in Bashivan et al. \cite[pp. 7-8]{learning_eeg_repr}, Support Vector Machines, Random Forest \cite{random_forests}, and sparse Logistic Regression models were trained, using a multiple parameter setting.
\subsubsection{Data Preparation}
For the baseline methods, the data was preprocessed in a more simplified way than for the Neural Network models. This is due to the fact that the models used do not scale as well for high dimensional data, as opposed to the Convolutional Layers in the Neural Network models. Features are thus computed for the whole sequence, rather than extracting them per certain time steps. This results in data that is no longer sequential in nature, but describes the whole observation.
For each of the 19 channels available in the data, Fast Fourier Transform was applied on the whole length of a channel's data. The magnitude of each band (theta, alpha, and beta) was then extracted and stored as the channel's features. In addition, an average over all channels was computed. The resulting feature vector is then of size 60: \textit{(19 channels + additional averaged channel) $\times$ 3 frequency bands}.
\subsubsection{Models}

\subparagraph{Support Vector Machines}
A Support Vector Machine classifier constructs a hyperplane (or set of hyperplanes) in the data space, which should separate the different classes. A model was trained for each combination of the following parameters:
\begin{itemize}
	\item The penalty parameter \textbf{C} - from range of \{0.01, 0.1, 1, 10, 100\}
	\item The $\gamma$ kernel coefficient - from range of \{0.1, 0.2, .., 1, 2, .., 10\}
\end{itemize}

\subparagraph{Random Forests}
An ensemble classification method. A number of Decision Trees are fit on different subsets of features of the training data. When classifying a data sample, the class which is given the majority of votes among the Decision Trees is chosen. A model was trained for each of the following numbers of Decision Trees: 5, 10, 20, 50, 100, 500, 1000.

\subparagraph{Logistic Regression}
A model was trained, using the $l_1$ as the norm for penalization, for each of the following penalization parameters \textbf{C}: \{0.01, 0.1, 1, 10, 100, 1000\}.

@@

The Neural Network models were trained on minibatches of size 32, using Adam \cite{Adam} as the optimization method.

@@

\begin{tabular}{ c|c } 
 Model & Accuracy \\ 

@@

 SVM & 7.8\% $\pm \, 4.1\% $\\
 Random Forest & 36.7\% $\pm \, 7.3\%$ \\
 Logistic Regression & 33.6\% $\pm \, 5.8\%$ \\

@@

 2 layer unidirectional LSTM & 18.6\% $\pm \, 2.0\% $\\
 2 layer bidirectional LSTM & 83.7\% $\pm \, 4.1\%$ \\
 3 layer bidirectional LSTM & 83.4\% $\pm \, 3.0\%$ \\
 5 layer bidirectional LSTM & 75.8\% $\pm \, 5.6\%$ \\
 2 layer bidirectional LSTM with 0.1 dropout & 84.1\% $\pm \, 4.2\%$ \\
 2 layer bidirectional LSTM with 0.2 dropout & 79.9\% $\pm \, 5.0\%$ \\

@@

 2 layer bidirectional LSTM with 0.1 dropout & 89.2\% $\pm \, 4.4\%$ \\
